---
title: "Associations of variates with mutual information"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Associations of variates with mutual information}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
    dev='png',
    fig.asp = 1/sqrt(2),
    fig.width = 8,
    out.width='100%',
    collapse = TRUE,
    comment = "#>"
)
options(knitr.kable.NA = '')
library(inferno)
## Tol colour-blind-friendly palette
palette(c(black='black',
        red='#EE6677',
        blue='#4477AA',
        green='#228833',
        yellow='#CCBB44',
        purple='#AA3377',
        cyan='#66CCEE',
        grey='#BBBBBB',
        midgrey='#888888'))
```

## Association and Mutual Information

A probability quantifies our degree of belief in the *values* of some variates, given that we know the *values* of some other variates. This belief also reflect our uncertainty: maximal for a belief of 50%, minimal for a belief of 0% or 100%.

In some cases we wish to quantify our uncertainty *in an average sense*, rather than about specific values, or given knowledge of specific values. In other words, we ask: on average, how uncertain are we about a set of variates, given that we have knowledge of another set? If this average uncertainty is low, then we say that the two sets of variates are highly *correlated* or *associated*; otherwise, that the are *uncorrelated*. Here we are using the word 'correlation' in its general sense, not in the specific sense of "linear correlation" or similar.

Traditionally the [Pearson correlation coefficient "$r$"](https://mathworld.wolfram.com/CorrelationCoefficient.html) is used to quantify association. It is extremely limited, however. It is essentially [based on the assumption that all variates involved have a joint Gaussian distribution](https://doi.org/10.1080/01621459.1954.10501231). As a consequence, it is a measure of *linear* association, rather than general association. For instance, if the graph of two continuous variates $X$ and $Y$ is a semicircle, then it means that $Y$ is a function of $X$, and is therefore perfectly associated with $X$: if we know $X$, then we can perfectly predict the value of $Y$. Yet the Pearson correlation coefficient between the two variates is $0$ in this case, simply because the functional dependence of $Y$ on $X$ is not linear:
```{r}
X <- seq(-1, 1, length.out = 200)
Y <- sqrt(1 - X^2)
r <- cor(X, Y, method = 'pearson')
plot(x = X, y = Y, type = 'p', main = paste0('Pearson correlation: ', signif(r, 2)))
```
Similar limitations of the Pearson correlation coefficient are demonstrated by the ["Anscombe quartet"](https://doi.org/10.2307/2682899) of datasets.



*(TO BE CONTINUED)*
